---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction
This paper aims to drive more competition and reduce collusion in the Canadian grocery sector. The same research can be done to other sectors to achieve similar goals like increasing competition and decreasing collusion in the sector. We use SQL @SQL to clean the raw data and R @citeR to analyze and visualize cleaned data. We create three figures @ggplot to show the distribution and relation between current price and old price of products from different grocery vendors. We found that the grocery vendors tend to use competitive pricing tactics or responses to market demand dynamics due to many items have remained the same price or become cheaper.revealing how different vendors position themselves in the market and how prices evolve over time.

# Data {#sec-data}

## Overview

All datas aim to foster competition and reduce market collusion in the Canadian grocery sector by assembling a database of historical grocery prices from prominent vendors. This dataset serves academic and legal purposes, enhancing transparency in grocery pricing strategies.The data encompasses prices from eight major grocery vendors in Canada, including Voila, TandT, Loblaws, and others. Prices are specifically for "in-store pickup" in Toronto, collected via screen-scraping of vendor websites. This method captures customer-facing prices but may miss details available in internal APIs. Below is the description for the variable we are taking to research:

- vendor: One of the 7 grocery vendors.
- current_price: Price at time of extract.
- old_price: An "old" struck-out price. This indicates that there was a Sale on. This is how you can differentiate a price-drop advertised as a Sale vs. a "quiet" decrease without a Sale.
- units: Units (grams, kg, number of items in package). May be blank for some vendors/products.

The dataset facilitates analysis through visualizations like price trends over time for products such as Becel Margarine and Sourdough Bread. Each visualization aids in comparing pricing strategies across different vendors and times. Exploratory data analysis provides insights into pricing patterns, with summary statistics highlighting average prices, standard deviations, and price ranges. Comparative analysis across vendors or periods can spotlight trends or anomalies in pricing behaviors.

## Measurement
	
The measurement process involves converting observed grocery pricing phenomena into a structured dataset. This transformation starts with the collection of price data from various Canadian grocery vendors' websites through screen-scraping techniques. Each price observation captures the 'current price' at the time of data collection and the 'old price', which indicates the previous price point before any discounts or changes. This approach allows us to trace price fluctuations over time and analyze market behaviors.

To ensure accuracy and relevance, the data is collected for items available for in-store pickup in a specific Toronto neighborhood, providing a consistent geographical context. The 'units' associated with each product, such as grams or kilograms, are noted to standardize price comparisons. This meticulous approach to data measurement facilitates a comprehensive analysis of pricing strategies across different vendors, revealing trends and potential collusive behaviors within the market.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.








# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


